Validation of Visual Statistical Inference, Applied to Linear Models

### 2013-03-20 paper_jasa.tex
expected power curve changed

\noindent Ideally, the visual test statistic is defined and constructed using the grammar of graphics \citep{wilkinson:1999,hadley:2009}, consisting of type, and specification of aesthetics, necessary for complete reproducibility. \\
\noindent
The visual test statistic is compared with values $T(y_0)$ using a lineup, which is defined as:

@@

\begin{dfn}\label{dfn:pvalue}
The $p$-value of a lineup of size $m$ evaluated by $K$ observers is  given as

\begin{equation}\label{binom}
P(X \ge x) = 1 - \text{Binom}_{{K, 1/m}} (x-1) = \sum_{i=x}^{K} {K \choose i} \left(\frac{1}{m}\right)^{i} \left(\frac{m-1}{m}\right)^{K-i}
\end{equation}

@@

\noindent An important difference between conventional and visual testing is that lineups will depend on observers evaluation. Thus $X$, the number of observers who identify the actual data plot from the lineup, affects the estimation of power  and the power is estimated by
\[
\widehat{Power}_{V} (\theta) = {Power}_{V, K} (\theta) = 1 - F_{X, \theta} (x_{\alpha} - 1),
\] 
Here $F_{X, \theta}$ is the distribution of $X$ and $x_\alpha$ is such that $P(X \ge x_{\alpha}) \le \alpha$. Note that the distribution  $F_X$ depends on which hypothesis is true:
under the null hypothesis, $X \sim Binom_{K, 1/m}$, leading to:
\[
Power_V(\theta, K)= 1 - Binom_{K, 1/m} (x_\alpha - 1),
\]
%where $x_\alpha$ is such that $P(X \ge x_{\alpha}) \le \alpha$. 
If the alternative hypothesis is true, with a fixed parameter value $\theta$, we can assume that an individual's probability to identify the data plot depends on the parameter value, and $X_i \sim Binom_{1, p_i(\theta)}$. Assessing an individual's skill to identify the actual data plot will require that an individual evaluates multiple lineups.

%\noindent Note that $F_{X, \theta}$ is the distribution of $X$ under the alternative with given parameter value $\theta$.

%\red{there may be a flow in this definition. Due to the discrete nature of binomial (specially $x_{\alpha}$) this definition would only give a conservative lower limit of the actual power. I feel like we should not define power using binomial.}

@@

%\green{Notation changes $i$ to $j$, and $p$ to $\pi$. Can someone please decide which to use, and make the appropriate changes? How does the ``Let $X_{j\ell} \sim B_{1, \pi_{j\ell}}$...'' below relate to the ``let $X_i$ be a binary random variable...'' above? It looks like this is a repeat. One needs to go once notation is clarified.}

%\red{Here is what my suggestion for notation \\  Use $B$ for logistic model and $\beta$ for linear model \\ Use $X$ to indicate logistic model covariate matrix and $U$ to denote binomial variable \\ $Y$ to indicate actual data. \\ Use $\pi_i$ instead of $p_i$ since $p_i$ is used for $p$-value.  \\ Should I change everything accordingly? I see that we have been careful not to use $\beta$ for power! }

\subsection{Subject-specific abilities}~\label{sec:model}
%Calculating the power  is intricately related to the probability to identify the actual data plot from the lineup.

@@

%Now assume that the same lineup plot is evaluated by $K$ independent observers. %W.l.o.g.\green{Can we get rid of cliches like WLOG?} 
%Let the last plot be the data plot and plots 1 through $m-1$ the null plots. Let $n_i$ be the number of observers who picked plot $i$ as the plot with the structure least consistent with the null hypothesis, corresponding to random variable $N_i$.  Then  $(N_1, N_2, ..., N_m) \sim$ Mult$_{\pi_1, ..., \pi_m}$ with $\sum_i \pi_i = 1$ and $\sum_{k=1}^{m} n_k = K$. We can estimate $\pi_i$ as $\widehat{\pi_i} = n_i/K$. 
%
%For the distribution of $N_m$, %the number of times out of $K$ that observers picked the last plot to show the structure least consistent with the null hypothesis, 
%we get
%\[
%N_m\sim \left \{ 
%\begin{array}{ll}
%B_{K, \hh{E\left[ (1 - p_D)^{m-1}\right]}} & \text { under } H_1\\
%B_{K, 1/m} & \text { under } H_0
%\end{array}
%\right .
%\]
%
%By matching the expected values, we therefore get an expression  for the $p$-value $p_D$ in the data 
%%under $H_1$  %% no such thing aas $p$-value under H_1 - even though it's just an unfortunate choice of word.
%as:
%\begin{equation}\label{eqn:power_estimate}
%%{E[p_D]} = 1- \left(\frac{E[N_m]}{K}\right)^{1/(m-1)}.
%\hh{1 - E[N_m]/K = (m-1)E[p_D] + O(E[p_D^2])}
%\end{equation}
%
%Note that the left hand side in Equation \ref{eqn:power_estimate} is independent of the parameter $\theta$, but just based on the lineup evaluation by independent observers. While this allows us to compute this value independently of the parametric setting, it also means that observers might pick out the data plot for a reason not related to the value of $\theta$. 
%\hh{Also note, that this equation yields an estimate of $p_D$ that is bounded by $1/(m-1)$. We should therefore not make the decision of significance depend on this estimate, but only use this equation after the fact. 
%Both of these topics are further highlighted in the application section.}

%Picking out the data plot is still a measure of how different the data plot is compared to the null plots, and therefore is a test statistic of the visual test.

%This allows us to compute this value independently of the parametric setting. We therefore define the {\it signal strength} of a  visual test as the following:
%
%\begin{dfn} \label{dfn:signal}
%The {\it signal strength} $\sigma_T$ of a visual test is defined as the probability to identify the data plot.
%\end{dfn}
%
%In a lineup of size $m$, evaluated by $K$ independent observers,  we can estimate the signal strength of the test as
%\[
%\sigma_T^{m-1} = \frac{n_m}{K}.
%\]
%
%In a parametric setting, signal strength of a test coincides with the power of the test.

%Theoretical power for the regression parameters is derived in the next section.
%\subsection{Estimating Power from Lineups}
%
%We estimate the empirical power from responses on a specific lineup plot  generated with known values of sample size ($n$), variance ($\sigma^2$) and regression parameter ($\beta$) in model \ref{multi}. Suppose, we have responses from $K$ independent observers with $u$ identifications of the actual plot. This gives an estimated power of
%\begin{equation}\label{power_est} Power(\beta)=\frac{u}{K} \hspace{1cm} 0 \le u \le K \end{equation}

%From model \ref{eqn:mixed} we obtain %$\pi=Pr(Y=1|X_1, X_2, ..., X_p)$, 
%the power of the underlying testing procedure as a population average %by ignoring the random effect part. 
%%Thus we have the overall probability of success ($\pi$ or power) 
%for specified sample size ($n$) and  variance ($\sigma$)  as 
%
%\begin{equation}\label{eqn:power} 
%Power(\beta) = \pi=Pr(Y=1|\beta, n, \sigma) 
%\end{equation}
%
%The main covariate factors that determine the power of the test are sample size ($n$), $\sigma$ and the parameter $\beta$. combining these we obtain one single covariate effect denoted by $E$ and defined as $E = \beta/ \sigma*\sqrt{n}$. Using equation \eqref{eqn:mixed}, we then have the power as
%
%\begin{equation}\label{eqn:effect} 
%Power(E) = \pi=Pr(Y=1| E) 
%\end{equation}

@@

%Theorem \ref{thm:power} helps to compute the expected power of visual inference.
%\begin{thm}\label{thm:power}
% equation \eqref{power_exp}  yields expected power of visual inference as Power($\beta$) = $E(1-p_D)^{m-1}$ .
%\end{thm}
% 
%{ \em proof}
%
%Since $p_0 \sim \text{Beta}(1,m-1)$,  for $t \in (0,1)$ we have the distribution function of $p_0$ be
%\begin{eqnarray*}
%F{p_0}(t) &=& (m-1) \int_0^t{(1-p_0)^{(m-2)}dp_0} \\
%  &=&  -(m-1) \int_1^{1-t}{x^{(m-2)}dx}\\
%  &=& \left[ x^{m-1}\right]^1_{1-t}\\
%  &=&1-(1-t)^{m-1} \rightarrow 1 \quad \text{as} \quad m \rightarrow \infty
%\end{eqnarray*}
%
%% verified this computation from http://www.wolframalpha.com using command int (m-1)*(1-p)^(m-2) dp, p=0,t
%
%Thus expected power in equation \eqref{power_exp} be
%\begin{eqnarray*}
% Pr(p_{B} < p_0) &=& 1- Pr( p_0 \le p_{B} ) \\
%  &=& 1- \int_{0}^{1}{Pr( p_0 \le p_{B} |p_{B} =t) f_{p_{B} }(t)dt } \\
%  &=& 1-  \int_{0}^{1}{F{p_0}(t) f_{p_{B} }(t)dt } \\
%  &=& E(1-p_D)^{m-1}   \\
%%  & \rightarrow &1-  \int_{0}^{1}{ f_{p_{B} }(t)dt }  = F_{p_D}(0) \quad \text{as} \quad m \rightarrow \infty
%\end{eqnarray*}
%
%
\begin{figure}[hbtp]
   \centering
       \scalebox{.6}{\includegraphics{power_expected_k.pdf}}
       \caption{Comparison of the expected power of the visual test for different $K$ (number of observers) with the power of the conventional test, for $n =100$, $\sigma = 12$ and $m$=20.  
       \hh{Could you standardize the $x$ axis to $\beta/\sigma$? That will give a better insight to how close we need to be to the null hypothesis to have visual power dip below conventional power.}}
       \label{fig:power_expected}
\end{figure}

@@

After many small pilot studies with local personnel, it was clear that some care was needed to set up the human subjects experiments. It was best for a observer or a subject to see a block of 10 lineups with varying difficulty, with a reasonable number of ``easy'' lineups. The explanations about each experiment (below) includes an explanation of how the lineups were sampled and provided to the subjects, for each of the three experiments.

@@

The number of evaluations required for each lineup to provide reasonable estimates of the \blue{ proportion correct ($\hat p$) is determined by the variance of $p$}. Suppose $\gamma$ denotes the conventional test power for each parameter combination shown in Table \ref{tbl:experiment_params}. Since the expected power of visual inference is very close to the power of conventional test (Figure \ref{fig:power_expected} with $K=1$) \blue{we consider $\gamma =p$}.
For a given proportion $\gamma$ it is desired to have a margin of error (ME) less than or equal to 0.05. Thus we have $ME =1.96 \sqrt{ \gamma(1-\gamma) / n_{\gamma} } \le 0.05$ which gives us the estimation of minimum number of evaluations $$n_{\gamma} \geq \frac{\gamma(1-\gamma)}{(0.05/1.96)^2}.$$  
% \hh{we're not even close to $n_\gamma$ with ten lineups .... we need more in the order of 400. Let's not emphasize this so much.}

@@

\begin{table}[hbtp]
\caption{Combination of parameter values, $\beta_2$,  $n$ and $\sigma$, used for each of the simulation experiments.} % title name of the table
\centering
\scalebox{.9}{
\begin{tabular}{c c c c c}
% creating 10 columns
\hline\hline

@@

\begin{table}[hbtp]
\caption{Number of subjects, gender and total lineups seen, and number of distinct lineups for all three experimental data sets. Note that in some of the lines the number of male and female participants does not add up to the total number of participants due to missing demographic information. }%\blue{Notice that we do not have data on one lineup (so we have only 29 lineups) for experiment 3 due to the screening process.}}
\begin{center}
\begin{tabular}{ccrrcc}
   \hline \hline
 Experiment & Subject & Male & Female & Responses &Lineup\\ 
    \hline
1 & 239 & 121 & 107 & 2249 &  60 \\ 
  2 & 351 & 185 & 164 & 3636 &  70 \\ 
  3 & 155 & 103 &  52 & 1511 &  29 \\  
   \hline
\end{tabular}
\end{center}
\label{tbl:summary}
\end{table}

@@

For experiment 3 we do fit intercepts: both a fixed  and subject-specific random effects, since forcing power
%While for experiment 1 and 2 we don't allow intercept but offset, for experiment 3 an intercept is estimated to allow the power curves to go above 0.05 when $E=0$. This is because forcing power 
to be fixed at 0.05 for $E=0$ is not justified by the data. % (Figure \ref{fig:power_loess_effect}). *** You cannot reference the figure before you explain it!!!

%For all the three experiments, each subject saw a lineup just once. This does not allow fitting a lineup random effect, in addition to a subject random effect, but it does simplify the model since $\tau_{\ell i} = \tau_i$.

%The subject-specific probability, $p_i$, of successful evaluation of a lineup is then estimated using Equation \ref{eqn:mixed_power}. 
For computation we use package {\tt lme4} \citep{lme4:2011} and software R 2.15.0 \citep{R}. $p$-value calculations are based on asymptotic normality.

Table \ref{tbl:model_par} shows the parameter estimates of the mixed effects model of the subject-specific variation. The fixed effects estimates indicate that for all experiments  the proportion of correct responses increases as the effect increases. This effect is less pronounced  for experiment 3. The subject-specific variability is smaller for experiment 1, and relatively large for experiment 3. %The larger variability in the response data for experiment 3% as we see in Figure \ref{fig:power_loess_effect}  *** You cannot reference the figure before you explain it!!!
%contributes to the random effect variance estimate be much larger compared to those of other experiments. For the same reason we see the standard error for intercept estimate is much higher as well. Also notice the slope estimate for experiment three is lowest as well. It is because for small parameter values the power is much higher for experiment 3.  % \green{What do we learn about the models from these parameter estimates? Explain this here too.}

%\green{This section needs to have sufficient detail of the model fitting, output and diagnostics to feed into the next few sections discussing the findings.}

@@

Results for experiment 3 are quite different. This is the situation where we expect to see the potential of visual inference, and indeed we do: the power of visual inference is always high, and much higher than the conventional test at small effect sizes. There is no actual conventional power in this situation, because assumptions are violated.
The dashed line shows conventional power based on uncontaminated data, whereas the cross marks show effective power based on the coefficient estimated from the contaminated data.
% It is computed assuming that there is no contamination, using the non-contaminated part of the simulation model. The  conventional power of true effect of the contaminated data is computed anyway and presented in figure \ref{fig:power_loess_effect} by cross marks. For this we obtain the estimated slope for the contaminated data and compute the estimated effect for which conventional power is calculated. 
% 
%Note that the power structure is completely broken for contaminated data. %For smaller original effect we get huge power and that explains why we see so many small $p$-values for experiment 3

@@

%(scatter plot) is better than what we observed for experiment 1 (box plot). For experiment 3, observed power is always above the conventional test power since for contaminated data we can visually see a trend while conventional test fail to address that.

%The power curves obtained by fitting model \ref{eqn:mixed} to the data for all the experiments applying different screening criteria is shown in figure \ref{fig:power_screening}. This shows how the screening criteria may affect the results. The corresponding power curves of conventional test are also shown. Notice that no mater what criteria we apply the result does not change much for experiment 2. Also, estimated power curve with criteria 5 for experiment 3 is above the conventional test power curve.

%
%\begin{figure}[hbtp]
%   \centering
%       \scalebox{0.70}{\includegraphics{power_mixed.pdf}}
%       \caption{The overall power estimated from model \ref{eqn:mixed} are shown. The corresponding power curve for conventional test is shown for comparison.}
%       \label{fig:power_mixed}
%\end{figure}

%Each participant was shown a sequence of 10 lineup plots.  In total, 3629 lineups were evaluated by 324 people coming from many different locations across the globe.  The results of the experiment are summarized in Figure \ref{fig:power_observed} which shows the observed power from the survey data calculated using equation \eqref{eqn:power_estimate} along with 95\% confidence interval calculated using Fisher's exact method.

%\begin{figure*}[hbtp]
%   \centering
%       \scalebox{0.45}{\includegraphics{effect_power_exp1.pdf}}
%       \scalebox{0.45}{\includegraphics{effect_power_exp2.pdf}}
%       \caption{Observed power shown by loess smoother and the power of conventional test for different effect size $E$ as per experiment 1 and 2.}
%       \label{fig:power_effect1}
%\end{figure*}

%\begin{figure*}[hbtp]
%   \centering
%       \scalebox{0.45}{\includegraphics{effect_power_subject_exp1.pdf}}
%       \scalebox{0.45}{\includegraphics{effect_power_subject_exp2.pdf}}
%       \caption{Subject specific power obtained from equation \eqref{eqn:effect} and the power of UMP test for different effect size $E$ as per experiment 1 and 2.}
%       \label{fig:power_effect_subject1}
%\end{figure*}

%\begin{figure*}[hbtp]
%   \centering
%       \scalebox{0.4}{\includegraphics{power_loess_exp3.pdf}}
%       \caption{Observed power shown by loess smoother with simultaneous bootstrap confidence band and the power of conventional test for sample size $n= 100$ and $\sigma = 5$ as per experiment 3. Notice that for experiment 3, power does not seem to depend on the parameter $\beta$. \blue{Di/Heike, I am not sure how I can relate this to demonstrate the power of visual inference when normal conditions are not met. Does this really test our hypothesis $H_0: \beta=0$ or some other hypothesis about the existence of cluster in the data?}}
%       \label{fig:power_loess3}
%\end{figure*}

%\begin{figure}[hbtp]
%   \centering
%       \scalebox{0.40}{\includegraphics{power_model.pdf}}
%       \caption{Estimated power curve from equation \eqref{eqn:power} along with 95\% confidence interval for sample size $n$ = 100 and $\sigma$ = 12.  The corresponding power curve for Uniformly Most powerful (UMP) test is shown for comparison.}
%       \label{fig:power_model}
%\end{figure}

%We fit model \eqref{eqn:mixed} to the survey data obtained from the simulation experiment. The estimated overall power curve obtained from equation \eqref{eqn:power} is shown in Figure \ref{fig:power_model}. Model \ref{eqn:mixed} also gives the subject specific power curves shown in Figure \ref{fig:power_subject}. The plot includes 20 randomly selected subject-specific power curves. Notice that the power curve estimated for one subject is above the UMP test power curve.

%\begin{figure}[hbtp]
%   \centering
%       \scalebox{0.45}{\includegraphics{power_subject.pdf}}
%       \caption{Estimated subject specific power curve from model \ref{eqn:mixed} for sample size $n$ = 100 and $\sigma$ = 12.  The corresponding power curve for Uniformly Most powerful (UMP) test is shown for comparison.}
%       \label{fig:power_subject}
%\end{figure}

@@

%Different subjects are expected to have different skill levels in evaluating lineups. This is accounted for in the mixed effects model (Equation \ref{eqn:mixed}).   
\blue{Subject-specific proportion correct $\hat p_i(E)$ is obtained using Equation \ref{eqn:mixed_power} and it is used to obtain power for $K=5$}. Figure  \ref{fig:power_mixed_subject} shows power curves for both the overall experiment and subject-specific variations. The thick continuous line shows overall estimated power, the thinner lines correspond to subject-specific power curves. For comparison, the  dashed lines show power curves of the conventional test. Subject-specific power is quite different between the three experiments. In experiment 2 subjects performed similarly, and substantially better than the conventional test. In experiment 1 there is more variability between subjects, with some doing better than the conventional test on large effects. In experiment 3 there is the most subject-specific variation. Some subjects performed substantially better than the conventional test, and on average the visual test was better.

@@

%\green{third experiment doesn't really have a p-value, does it? Need to explain this. Why are there still so many small p-values in experiment 3?} \blue{$p$-values are computed fitting the full model to the contaminated data anyway. So many small p-values are coming from the fact that contamination makes the slope negatively significant. For example, for some slope the contamination reverses the slope (positive to negative) and negative slope got significant. It is like the effect of influential outlier.}

@@

From the experimental data the visual $p$-values are estimated based on Definition \ref{dfn:pvalue}. Figure \ref{fig:pval_definition} displays resulting estimates for each lineup against the conventional $p$-value.  The pattern of visual $p$-values is interesting: for small $p$-values the visual estimates tend to be very small, while lineups with larger $p$-values result in very large visual estimate, 
% If there is a signal in the data plot, visual $p$-value has a tendency to take a smaller value (small enough to reject $H_0$) and if not it takes a larger value (large enough to not reject $H_0$) 
% 
 giving a clear indication to reject $H_0$ or not. This is why we do not see lot of visual $p$-values  between 0.05 and 0.8 especially for experiment 2. This guides the researcher to make decision confidently while conventional $p$-values take some marginal $p$-values making the decision whether to reject or not harder.
 % for which researchers get confused whether to reject or 'not reject' $H_0$. 
 For visual tests this is not common.

\begin{figure*}[hbtp]
   \centering
       \scalebox{0.6}{\includegraphics{p_val_definition.pdf}}
       \caption{Conventional test $p$-value ($p_D$) vs visual $p$-value obtained from the definition . Values are shown on square root scale. }
    %   \hh{Mahbub, another change: could you change the y label to 'estimate of visual p-value $\hat{p_D}$ ?}
       \label{fig:pval_definition}
\end{figure*}

%We also see the same pattern from the loess smoother fitted through the points in Figure \ref{fig:pval_definition}. 
%The sigmoid shaped loess curves indicate a clear separation of visual $p$-values. For experiment 1, most of the visual $p$-values are above 0.4 when conventional $p$-values are above 0.2. This is more evident for experiment 2 where most of the visual $p$-values are above 0.8.

For experiment 3, we see that  the visual $p$ values are very small no matter what the conventional $p$-values are. This is expected since conventional test fails to reject $H_0$ where we should reject, whereas the visual test performs well.
% and the visual test does the job nicely.

%With multiple subjects it is possible to estimate the visual $p$-value, and a measure of plot signal strength. The visual $p$-value is calculated from the binomial formula (Equation \ref{binom}). Assuming $H_o$ is true, the probability of picking the actual data plot by chance is $1/m$, and the visual $p$-value is the probability that $x$ or more subjects select that plot. Figure \ref{fig:pval_definition} displays this visual $p$-value for each lineup against the conventional $p$value.

%Similary, the strength of the signal, is calculated using the proportion of subjects picking actual data plot. It is closely related to the power of the test. From Equation \ref{eqn:power_estimate}  we can write $1 - E[N_m]/K \approx (m-1)E[p_D] $. This gives the signal strength estimate as  
%\[
%\hat p_D= (1- {u}/{K})/({m-1})
%\]
%where $K$ is the total number of independent evaluations on a lineup and $u$ is the number of successful evaluations. Figure \ref{fig:pval_plot_signal} shows the plot signal strengths in relation with conventional test $p$-values, on a square root scale. For experiment 3, technically, there are no conventional $p$-values, but the $p$-values are computed in the same manner as experiment 2, assuming that the data satifies the assumptions. The blue line is a loess fit to the data, and the grey line indicates equality of the two axes. The relationship between the conventional $p$-value and the plot signal strength is very strong, particularly when the values are small. As the conventional $p$-value increases, the estimated signal strength increases -- subjects find it harder to detect the actual data plot at larger $p$-values. The rise is very sharp for experiment 1, which indicates that subjects have a harder time detecting the actual data plot for even small $p$-values.

%\green{Where is Figure \ref{fig:pval_definition} referred to and discussed?}
% \blue{Are we going to keep both figures \ref{fig:pval_definition} and \ref{fig:pval_plot_signal}?}
%
%\green{Explain why the term ``plot signal'' is introduced}

%, to assess the signal strength we examine the proportion of correct picks, with multiple users and the $p$-value of the actual data plot. For experiment 3 the \green{How is p-value calculated here?} Figure \ref{fig:pval_pcorrect} \blue{(should I bring this figure back?)}  displays scatterplots of these two variables for each experiment. As the $p$-value increases the proportion of correct responses falls, evidence of direct association between proportion of correct responses and conventional test $p$-values. When $p$-value is larger than 0.15, it is almost impossible to correctly identify the observed plot in the lineup. We observed this for both experiment 1 and 2. For experiment 3, the scenario is different, because there are two structures, the association between variables, and the cluster of contaminated points.% as in this case people might have observed something other than the effect $E$.

%Yet, it is possible to obtain an estimate the strength of the signal in the actual data plot, when there are multiple evaluations, using Lemma \ref{lemma}.

%Note that $\hat{p}_D$ is bounded within $(0, 1/(m-1))$. This estimate has to be interpreted carefully, since the choice of a large lineup size, $m$, will make the value look `significant' in a conventional sense. Values close to the upper bound of $1/(m-1)$ should therefore be interpreted as `at least as high', rather than their actual numeric value.
%\hh{For $p$-values below $1/(m-1)$ visual $p$-values are very similar to actual $p$-values, the correlation for those values is: 0.36, 0.79, and -0.48 based on $n_1 = 44, n_2 = 45$, and $n_3= 26$ lineups, respectively.  }  \blue{This is not happening as we see in figure \ref{fig:pval_plot_signal}. But it is true for experiment 2, right?. }
%is equivalent to conventional test $p$-value when $p$-value is less than or equal to $1/(m-1)$.  
%and beyond that the plot signal strength levels off on a point bigger than 0.05. Thus if plot signal strength is less than 0.05, we may decide to reject the null hypothesis.

%\begin{figure*}[hbtp]
%   \centering
%       \scalebox{0.5}{\includegraphics{p_val_plot_signal.pdf}}
%       \caption{Conventional test $p$-value ($p_D$) vs plot signal strength ($\hat p_D$). The solid line is a 45 degree line which represents the equity shown in Equation \ref{eqn:power_estimate}. p-values are shown on square root scale scale.}
%    %   \hh{Mahbub, another change: could you change the y label to 'estimate of visual p-value $\hat{p_D}$ ?}
%       \label{fig:pval_plot_signal}
%\end{figure*}

@@

%Each row corresponds to a different $\beta$ value ordered top to bottom from lowest to highest. Columns correspond to the other levels of the experiment, sample size $(n)$, error standard deviation $(\sigma)$ and replicate. (Empty cells indicate no lineup shown with this combination.)

@@

%\begin{figure*}[hbtp]
%   \centering
%      \includegraphics[width=0.8\textwidth]{p_val_log_counts-a.pdf}
%       \includegraphics[width=0.8\textwidth]{p_val_log_counts-b.pdf}
%       \caption{Relative frequency of plot picks compared to other plots in the lineup plotted against the $p$-value (on log$_{10}$ scale) of each plot for all of the 60 individual lineups for experiment 1. Red indicates the plot with the lowest $p$-value, and blue indicates the actual data plot, when it is different from that with the lowest $p$-value. The columns of both top and bottom panel are ordered according to effect size, rows within panels show replicates for each parameter combination.}
%       \label{fig:P-val_log}
%\end{figure*}

>>>>

\noindent Ideally, the visual test statistic is defined and constructed using the grammar of graphics \citep{wilkinson:1999,hadley:2009}, consisting of type, and specification of aesthetics, necessary for complete reproducibility. The visual test statistic is compared with values $T(y_0)$ using a lineup, which is defined as:

@@

\begin{dfn}\label{dfn:pvalue}
The $p$-value of a lineup of size $m$ evaluated by $K$ observers is  given as 
$$
P(X \ge x) = 1 - \text{Binom}_{{K, 1/m}} (x-1) = \sum_{i=x}^{K} {K \choose i} \left(\frac{1}{m}\right)^{i} \left(\frac{m-1}{m}\right)^{K-i}
$$

@@

\noindent An important difference between conventional and visual testing is that lineups will depend on observers' evaluation. Thus $X$, the number of observers who identify the actual data plot from the lineup, affects the estimation of power  and the power is estimated by
\[
\widehat{Power}_{V} (\theta) = {Power}_{V, K} (\theta) = 1 - F_{X, \theta} (x_{\alpha} - 1),
\] 
Here $F_{X, \theta}$ is the distribution of $X$ and $x_\alpha$ is such that $P(X \ge x_{\alpha}) \le \alpha$. Note that the distribution  $F_X$ depends on which hypothesis is true:
under the null hypothesis, $X \sim Binom_{K, 1/m}$, leading to:
\[
Power_V(\theta, K)= 1 - Binom_{K, 1/m} (x_\alpha - 1),
\]
%where $x_\alpha$ is such that $P(X \ge x_{\alpha}) \le \alpha$. 
If the alternative hypothesis is true, with a fixed parameter value $\theta$, we can assume that an individual's probability to identify the data plot depends on the parameter value, and $X_i \sim Binom_{1, p_i(\theta)}$. Assessing an individual's skill to identify the actual data plot will require that an individual evaluates multiple lineups.

@@

\subsection{Subject-specific abilities}~\label{sec:model}

@@

@@

%Theorem \ref{thm:power} helps to compute the expected power of visual inference.
%\begin{thm}\label{thm:power}
% equation \eqref{power_exp}  yields expected power of visual inference as Power($\beta$) = $E(1-p_D)^{m-1}$ .
%\end{thm}
% 
%{ \em proof}
%
%Since $p_0 \sim \text{Beta}(1,m-1)$,  for $t \in (0,1)$ we have the distribution function of $p_0$ be
%\begin{eqnarray*}
%F{p_0}(t) &=& (m-1) \int_0^t{(1-p_0)^{(m-2)}dp_0} \\
%  &=&  -(m-1) \int_1^{1-t}{x^{(m-2)}dx}\\
%  &=& \left[ x^{m-1}\right]^1_{1-t}\\
%  &=&1-(1-t)^{m-1} \rightarrow 1 \quad \text{as} \quad m \rightarrow \infty
%\end{eqnarray*}
%
%% verified this computation from http://www.wolframalpha.com using command int (m-1)*(1-p)^(m-2) dp, p=0,t
%
%Thus expected power in equation \eqref{power_exp} be
%\begin{eqnarray*}
% Pr(p_{B} < p_0) &=& 1- Pr( p_0 \le p_{B} ) \\
%  &=& 1- \int_{0}^{1}{Pr( p_0 \le p_{B} |p_{B} =t) f_{p_{B} }(t)dt } \\
%  &=& 1-  \int_{0}^{1}{F{p_0}(t) f_{p_{B} }(t)dt } \\
%  &=& E(1-p_D)^{m-1}   \\
%%  & \rightarrow &1-  \int_{0}^{1}{ f_{p_{B} }(t)dt }  = F_{p_D}(0) \quad \text{as} \quad m \rightarrow \infty
%\end{eqnarray*}
%
%
\begin{figure}[hbtp]
   \centering
       \scalebox{.6}{\includegraphics{power_expected_k.pdf}}
       \caption{Comparison of the expected power of the visual test for different $K$ (number of observers) with the power of the conventional test, for $n =100$, $\sigma = 12$ and $m$=20.  
       }
       \label{fig:power_expected}
\end{figure}

@@

After many small pilot studies with local personnel, it was clear that some care was needed to set up the human subjects experiments. It was best for a observer or a subject to see a block of 10 lineups with varying difficulty, with a reasonable number of ``easy'' lineups. The explanations about each experiment (below) includes an explanation of how the lineups were sampled and provided to the subjects.

@@

The number of evaluations required for each lineup to provide reasonable estimates of the \blue{ proportion correct ($\hat p$) is determined by the variance of the number of correct evaluations}. Suppose $\gamma$ denotes the conventional test power for each parameter combination shown in Table \ref{tbl:experiment_params}. Since the expected power of visual inference is very close to the power of conventional test (Figure \ref{fig:power_expected} with $K=1$) \blue{we consider $\gamma =p$}.
For a given proportion $\gamma$ it is desired to have a margin of error (ME) less than or equal to 0.05. Thus we have $ME =1.96 \sqrt{ \gamma(1-\gamma) / n_{\gamma} } \le 0.05$ which gives us the estimation of minimum number of evaluations $$n_{\gamma} \geq \frac{\gamma(1-\gamma)}{(0.05/1.96)^2}.$$  
% \hh{we're not even close to $n_\gamma$ with ten lineups .... we need more in the order of 400. Let's not emphasize this so much.}

@@

\begin{table}[hbtp]
\caption{Combination of parameter values, $\beta_2$,  $n$ and $\sigma$, used for the simulation experiments.} % title name of the table
\centering
\scalebox{.9}{
\begin{tabular}{c c c c c}
% creating 10 columns
\hline\hline

@@

\begin{table}[hbtp]
\caption{Number of subjects, gender and total lineups seen, and number of distinct lineups for all three experimental data sets. Note that in some of the lineups the number of male and female participants does not add up to the total number of participants due to missing demographic information. }%\blue{Notice that we do not have data on one lineup (so we have only 29 lineups) for experiment 3 due to the screening process.}}
\begin{center}
\begin{tabular}{ccrrcc}
   \hline \hline
 Experiment & Subject & Male & Female & Responses &Lineup\\ 
    \hline
1 & 239 & 121 & 107 & 2249 &  60 \\ 
  2 & 351 & 185 & 164 & 3636 &  70 \\ 
  3 & 155 & 103 &  52 & 1511 &  29 \\  
   \hline
\end{tabular}
\end{center}
\label{tbl:summary}
\end{table}

@@

For experiment 3 we do fit intercepts: both a fixed  and subject-specific random effects, since forcing power to be fixed at 0.05 for $E=0$ is not \blue{ required by the experimental design}.

For computation we use package {\tt lme4} \citep{lme4:2011} and software R 2.15.0 \citep{R}. $p$-value calculations are based on asymptotic normality.

Table \ref{tbl:model_par} shows the parameter estimates of the mixed effects model of the subject-specific variation. The fixed effects estimates indicate that for all experiments  the proportion of correct responses increases as the effect increases. This effect is less pronounced  for experiment 3. The subject-specific variability is smaller for experiment 1, and relatively large for experiment 3.

@@

Results for experiment 3 are quite different. This is the situation where we expect to see the potential of visual inference, and indeed we do: the power of visual inference is always high, and much higher than the conventional test at small effect sizes. There is no actual conventional power in this situation, because assumptions are violated.
The dashed line shows conventional power based on uncontaminated data, whereas the cross marks show effective power based on the coefficient estimated from the contaminated data.

@@

@@

\blue{Subject-specific proportion correct $\hat p_i(E)$ is obtained using Equation \ref{eqn:mixed_power} and it is used to obtain power for $K=5$}. Figure  \ref{fig:power_mixed_subject} shows power curves for both the overall experiment and subject-specific variations. The thick continuous line shows overall estimated power, the thinner lines correspond to subject-specific power curves. For comparison, the  dashed lines show power curves of the conventional test. Subject-specific power is quite different between the three experiments. In experiment 2 subjects performed similarly, and substantially better than the conventional test. In experiment 1 there is more variability between subjects, with some doing better than the conventional test on large effects. In experiment 3 there is the most subject-specific variation. Some subjects performed substantially better than the conventional test, and on average the visual test was better.

@@

@@

From the experimental data the visual $p$-values are estimated based on Definition \ref{dfn:pvalue}. Figure \ref{fig:pval_definition} displays resulting estimates for each lineup against the conventional $p$-value.  The pattern of visual $p$-values is interesting: for small $p$-values the visual estimates tend to be very small, while lineups with larger $p$-values result in very large visual estimate,

 giving a clear indication to reject $H_0$ or not. This is why we do not see lot of visual $p$-values  between 0.05 and 0.8 especially for experiment 2. This guides the researcher to make decision confidently while conventional $p$-values take some marginal $p$-values making the decision whether to reject or not harder.

 For visual tests this is not common.

\begin{figure*}[hbtp]
   \centering
       \scalebox{0.6}{\includegraphics{p_val_definition.pdf}}
       \caption{Conventional test $p$-value ($p_D$) vs visual $p$-value obtained from the definition . Values are shown on square root scale. }
       \label{fig:pval_definition}
\end{figure*}

For experiment 3, we see that  the visual $p$ values are very small no matter what the conventional $p$-values are. This is expected since conventional test fails to reject $H_0$ where we should reject, whereas the visual test performs well.

@@

@@
